---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r readme-01, echo = FALSE}
knitr::opts_chunk$set(
  fig.asp = 2/3,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# learnrbook #

[![cran version](https://www.r-pkg.org/badges/version/learnrbook)](https://cran.r-project.org/package=learnrbook)

## Purpose

Package '**learnrbook**' comtains data in R objects and in foreign files used as examples in the
book "Learn R: As a Language" by Pedro J. Aphalo, New York: Chapman and Hall/CRC, 2020, ISBN 9780367182533".  

In addition this package exports as character vectors the names of the
packages used in the book, so as to make it easier to install them.

## Installation

Installation of the most recent stable version from CRAN:

```{r cran-instalaltion, eval=FALSE}
install.packages("ggpmisc")
```

Installation of the current unstable version from Bitbucket:

```{r bb-instalaltion, eval=FALSE}
# install.packages("devtools")
devtools::install_bitbucket("aphalo/ggpmisc")
```

## Documentation 

HTML documentation is available at (https://docs.r4photobiology.info/learnrbook/).

News about updates are regularly posted at (https://www.r4photobiology.info/).

## Contributing

Please report bugs and request new features at (https://bitbucket.org/aphalo/learnrbook/issues). Pull requests are welcome at (https://bitbucket.org/aphalo/learnrbook).

## Citation

If you use this package to produce scientific or commercial publications, please cite according to:

```{r}
citation("learnrbook")
```

## License

© 2018-2020 Pedro J. Aphalo (pedro.aphalo@helsinki.fi). Released under the GPL, version 2 or greater. This software carries no warranty of any kind.
