---
title: "Installation and use"
subtitle: "`learnrbook` `r packageVersion('learnrbook')`"
author: "Pedro  J. Aphalo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation and use}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

This package contains data sets used in the code examples in the book "Learn R 
...as you learnt your mother tongue" by Pedro J. Aphalo, available at
<https://leanpub.com/learnr>. The book uses mostly data sets included in the
R distribution itself and in the packages available through CRAN that are
described in the book, so few data sets are included as R data in this package. 
However, to demonstrate the reading of foreign data
formats I use in the book several files. These files are part of this
package and installed in the `extdata` folder.

## Installation of packages

Installation of the package itself is not different to that of any other
package available through CRAN.

```{r, eval=FALSE}
install.packages("learnrbook")
```

To install the current development version from Bitbucket (possibly buggy
or matching a revision of the book not yet published) use the code in the next
chunck. (Unless package 'devtools'
is already installed in the computer you are using, you will need to install
this package by uncommenting and executing the first line of the code below.)

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_bitbucket("aphalo/learnr-pkg")
```

In the code examples in the book I use more than 40 packages available from 
CRAN, and three not yet in CRAN but available from Github. They are listed at 
the start of each chapter. To facilitate their
installation and use, vectors containing package names used in each chapter of
the book are provided, as well as a vector will all the names together.

To install all the packages used in the book you can use.

```{r, eval=FALSE}
install.packages(learnrbook::pkgs_all)
```

If you rather install only the packages used in a given chapter use, for example,
for theplotting with `ggplot' chapter.

```{r, eval=FALSE}
install.packages(learnrbook::pkgs_ch_ggplot)
```

Even though three of the packages are not in CRAN, they are all listed in the
vectors of names in the hope that in the future they will be submitted to CRAN.

An additional vector `pkgs_at_github` provides the paths to the Github repositories for these 
three packages.

These can be installed as follows.

```{r, eval=FALSE}
# install.packages("devtools")
for (pkg in pkgs_at_github) {
  devtools::install_github(pkg)
}
```

## Use of examples files

Example files are installed in folder `extdata` under the root folder of the
installed package. The location and path to where a package is installed is
dependent on the operating system and account settings, and the settings in
the `.Rprofile` file. However, R provides a simple way for querying this path to
where the example files are stored.

```{r, eval=FALSE}
system.file("extdata", package = "learnrbook")
```

We can also easily get a list of the available example files.

```{r}
path2examples <- system.file("extdata", package = "learnrbook")
list.files(path2examples)
```

The formats of these files are:

| File name "tag" | Format | other |
|:-----------------|:--------|:--------------------------------|
| .SYS            | Systat | proprietary, format depends on operating system |
| .tif            | TIFF   | Bitmap, variations exist in bit depth and compression used  |
| .sav            | SPSS   | proprietary            |
| .txt            | ASCII  | open                   |
| .nc             | NetCDF | open                   |
| .xlsx           | Excell | proprietary, modern, XML based      |

The files are used in Chapter "Storing and manipulating data with R" and in
Chapter "Plotting maps and images".

The sources of the data in these files, and a short descriptions of the data
follows. However, see the individual help pages for additional details on the data sets.

| File           | when saved | Data |Source |
|:---------------|:-----------|:-----|-----------|
| BIRCH1.SYS     | 1992       | Birch seeedlings | Aphalo and Rikala (2002) |
| Book1.xlsx     | 2017       | artificial data | Original |
| dahlia-uva.tif | 2017       | photograph | Original, P. J. Aphalo |
| dahlia-vis.tif | 2017       | photograph | Original, P. J. Aphalo |
| logger_1.txt   | 2016       | from Yocto Puce datalogger | Original, P. J. Aphalo |
| meteo-data.nc  | 2017       | meteorological | see help for source |
| my-data.sav    |        | artificial |       |
| my-data.xlsx   |        | artificial |       |
| pet_ltm.nc     | 2014       | meteorological | Data of Anders Lindfors |
| pevpr.sfc.mon.ltm.nc | ???? | ???? | ???? |
| thiamin.sav    | ????       |           |         |



